Reanalysis Step 2: Multiple Imputation

### 2025-08-30 reanalysis-2-multiple_imputation.qmd
perfecxt prediction issue

@@

  # Temporal VA variables
  "cohort_id",                 # Data indicator only - not used as predictor
  "baseline_VA_logMAR",        # Baseline VA
  "intermediate_VA_logMAR",    # Intermediate follow-up VA (for excluded cohort)
  "outcome_VA_logMAR",         # 2-year outcome (true for submitted, missing for excluded)
  # NOT outcome_va_vi_binary - derived from logMAR threshold

@@

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = dplyr::row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2),
                       labels = c("No_DR", "NPDR", "PDR"),
                       ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    cohort_id = factor(cohort_id),
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  ) |>
  # Fellow-eye variables (computed within patient clusters)
  dplyr::group_by(e_mrn_deidentified) |>
  dplyr::mutate(
    n_eyes = dplyr::n(),
    eye_index = dplyr::row_number(),
    fellow_baseline_VA = dplyr::if_else(n_eyes == 2, baseline_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_intermediate_VA = dplyr::if_else(n_eyes == 2, intermediate_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_outcome_VA = dplyr::if_else(n_eyes == 2, outcome_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_dr = dplyr::if_else(n_eyes == 2,
                               person_dr[3 - eye_index],
                               factor(NA, levels = levels(person_dr), ordered = TRUE))
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-eye_index)
```

@@

# Temporal VA variables
custom_methods["baseline_VA_logMAR"] <- "pmm"
custom_methods["intermediate_VA_logMAR"] <- "pmm"
custom_methods["outcome_VA_logMAR"] <- "pmm"
custom_methods["cohort_id"] <- ""  # never impute

@@

# IDs and fellow-eye variables: never impute
custom_methods["row_id"] <- ""
custom_methods["e_mrn_deidentified"] <- ""
for (fv in c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")) {
  if (fv %in% names(custom_methods)) custom_methods[fv] <- ""
}

@@

# Never impute IDs (row = 0) and never use them as predictors (col = 0)
for (idv in c("row_id", "e_mrn_deidentified")) {
  if (idv %in% rownames(pred_matrix)) pred_matrix[idv, ] <- 0
  if (idv %in% colnames(pred_matrix)) pred_matrix[, idv] <- 0
}

# Never impute but can be used as predictors: ever_lapse_binary, cohort_id
if ("ever_lapse_binary" %in% rownames(pred_matrix)) pred_matrix["ever_lapse_binary", ] <- 0
if ("cohort_id" %in% rownames(pred_matrix)) pred_matrix["cohort_id", ] <- 0

@@

  # Check if cohort_id might be blocking
  if ("cohort_id" %in% names(df_mice)) {
    cohort_excluded <- sum(df_mice$cohort_id == "excluded")
    log_message(paste0("  - Records with cohort_id='excluded': ", cohort_excluded))
  }
}

@@

# Add back variables not included in imputation or transformations (exact row match on row_id)
df_original_full <- df_for_imputation |>
  mutate(row_id = dplyr::row_number())

vars_in_imputed <- names(imputed_datasets[[1]])
vars_to_add <- setdiff(names(df_original_full), c(vars_in_imputed, "row_id"))

>>>>

::: {.callout-important}
## Key Change from Previous Version
`cohort_id` has been removed from the imputation model to avoid perfect prediction issues. The variable perfectly predicted outcome missingness (all 'excluded' had missing outcomes, all 'submitted' had observed outcomes), preventing successful imputation. The variable is preserved and added back after imputation for tracking purposes.
:::

@@

  # Temporal VA variables
  # NOTE: cohort_id REMOVED - it perfectly predicts outcome missingness
  "baseline_VA_logMAR",        # Baseline VA
  "intermediate_VA_logMAR",    # Intermediate follow-up VA
  "outcome_VA_logMAR",         # 2-year outcome
  # NOT outcome_va_vi_binary - derived from logMAR threshold

@@

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = dplyr::row_number(),
    # Store cohort_id separately for later use (not in imputation)
    cohort_id_original = cohort_id
  ) |>
  dplyr::select(row_id, cohort_id_original, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2),
                       labels = c("No_DR", "NPDR", "PDR"),
                       ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  ) |>
  # Fellow-eye variables (computed within patient clusters)
  dplyr::group_by(e_mrn_deidentified) |>
  dplyr::mutate(
    n_eyes = dplyr::n(),
    eye_index = dplyr::row_number(),
    fellow_baseline_VA = dplyr::if_else(n_eyes == 2, baseline_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_intermediate_VA = dplyr::if_else(n_eyes == 2, intermediate_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_outcome_VA = dplyr::if_else(n_eyes == 2, outcome_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_dr = dplyr::if_else(n_eyes == 2,
                               person_dr[3 - eye_index],
                               factor(NA, levels = levels(person_dr), ordered = TRUE))
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-eye_index)
```

@@

# Temporal VA variables
custom_methods["baseline_VA_logMAR"] <- "pmm"
custom_methods["intermediate_VA_logMAR"] <- "pmm"
custom_methods["outcome_VA_logMAR"] <- "pmm"
# cohort_id removed from dataset to avoid perfect prediction

@@

# IDs and fellow-eye variables: never impute
custom_methods["row_id"] <- ""
custom_methods["e_mrn_deidentified"] <- ""
custom_methods["cohort_id_original"] <- ""  # Stored for later, not imputed
for (fv in c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")) {
  if (fv %in% names(custom_methods)) custom_methods[fv] <- ""
}

@@

# Never impute IDs (row = 0) and never use them as predictors (col = 0)
for (idv in c("row_id", "e_mrn_deidentified", "cohort_id_original")) {
  if (idv %in% rownames(pred_matrix)) pred_matrix[idv, ] <- 0
  if (idv %in% colnames(pred_matrix)) pred_matrix[, idv] <- 0
}

# Never impute but can be used as predictors: ever_lapse_binary
if ("ever_lapse_binary" %in% rownames(pred_matrix)) pred_matrix["ever_lapse_binary", ] <- 0
# cohort_id removed entirely to avoid perfect prediction issues

@@

  # Note about cohort_id removal
  log_message("  - cohort_id removed from model to avoid perfect prediction")

  # Check proportion missing
  pct_missing <- round(100 * n_miss_outcome / nrow(df_mice), 1)
  log_message(paste0("  - Percent missing: ", pct_missing, "%"))
}

@@

# Add back cohort_id and other variables not included in imputation
df_original_full <- df_for_imputation |>
  mutate(row_id = dplyr::row_number())

# First add back cohort_id from cohort_id_original
imputed_datasets <- lapply(imputed_datasets, function(imp_data) {
  imp_data$cohort_id <- imp_data$cohort_id_original
  imp_data$cohort_id_original <- NULL  # Remove the temporary column
  imp_data
})

# Then add any other variables not included in imputation or transformations
vars_in_imputed <- names(imputed_datasets[[1]])
vars_to_add <- setdiff(names(df_original_full), c(vars_in_imputed, "row_id"))
